---
title: "Structural Equation Modeling (SEM)- Based Meta-Analysis"
author: "Mike Cheung"
date: last-modified
toc: true
format: 
  html: 
   code-fold: true
  # number-sections: true
#theme: #superhero #quartz
#df-print: paged
execute: 
 #echo: false
 cache: true
 freeze: false
editor: visual
---

```{r echo: false}
#| message: false
#| warning: false

library(metaSEM)
library(symSEM)
```

------------------------------------------------------------------------

## **Background** {.unnumbered}

::: callout-note
### Conceptualization of Meta-Analysis as SEM

-   Treats studies in a meta-analysis as subjects in SEM.

-   Observed effect sizes correspond to observed variables; population (true) effect sizes correspond to latent variables.

-   Mean and variance of effect sizes equate to average effect and heterogeneity variance in meta-analysis.

-   Sampling variance is analogous to measurement error in SEM.

-   Study characteristics (covariates) can predict effect sizes in cases of excess heterogeneity.

-   **Advantages**:

    -   Handle missing data via **Full Information Maximum Likelihood (FIML)**.\
    -   Fix known sampling variances using **definition variables**.\
    -   Visualize models with path diagrams (e.g., latent heterogeneity).
:::

------------------------------------------------------------------------

## **Computing Effect Sizes**

::: callout-note
### Definition of Effect Size

-   Represents the magnitude of a phenomenon in quantitative terms.
-   Can be standardized (e.g., Cohen’s d, correlation coefficient) or unstandardized (e.g., raw mean difference).
:::

-   The generic effect-size for each `study i` is de notated as $y_i$ and is represented as: $$y_i = f_i + e_i$$ $y_i$ = sample (observed) effect size for the ith study, $f_i$ = population (true) effect size, $e_i$ = error term in the observed effect-size, $var(e_i)$ = a.k.a $v_i$ sampling variance (usually known in MA)

#### Standardized Mean Difference (SMD)

**Equations**: $$
y_{\text{SMD}} = \frac{\bar{X}_1 - \bar{X}_2}{S_{\text{pooled}}}, \quad 
S_{\text{pooled}} = \sqrt{\frac{(n_1-1)S_1^2 + (n_2-1)S_2^2}{n_1 + n_2 - 2}}
$$

**Sampling Variance**: $$
v_{\text{SMD}} = \frac{n_1 + n_2}{n_1 n_2} + \frac{y_{\text{SMD}}^2}{2(n_1 + n_2)}
$$

**R Code**:

```{r}
compute_SMD <- function(m1, m2, sd1, sd2, n1, n2) {
  pooled_sd <- sqrt(((n1 - 1)*sd1^2 + (n2 - 1)*sd2^2) / (n1 + n2 - 2))
  smd <- (m1 - m2) / pooled_sd
  v_smd <- (n1 + n2)/(n1 * n2) + smd^2/(2*(n1 + n2))
  return(data.frame(y = smd, v = v_smd))
}

# Example: Compute SMD for two groups
compute_SMD(m1 = 10, m2 = 8, sd1 = 2, sd2 = 1.5, n1 = 50, n2 = 50)
```

------------------------------------------------------------------------

## Univariate approach

#### **Fixed-Effect Model**

-   The population effect-sizes across all studies are assumed to be equal or the same. In other words, there is a true effect (in the population, but it is the same for all individual study).
-   The differences in the observed effect-size is simply due to random (sampling) error attached to each individual studies with a variance of $v_i$ (assumed to be known).
    -   $v_i$ is assumed to be known.
-   The fixed-effect model focuses on the studies in the current meta-analysis with the same design, measures, and samples (i.e,. the findings can't be generalized beyond the studies in the meta analysis).

### Model Specification

**Equation**: $$
y_i = \mu + e_i, \quad e_i \sim N(0, v_i)
$$

**SEM Representation**:

-   No latent variables because $\mu_i$.\
-   Fixed parameter: $\mu$ (common effect). This is a.k.a the model implied mean -SEM (or average effect across studies- MA)
-   Known sampling variance $v_i$ not random but fixed (not estimated in the model).

::: callout-note
### Dataset

-   Data is from Jaramillo et al 2005
-   contains 61 subjects -SEM (61 studies- MA)
-   $y_i$ = observed variable- SEM (effect- MA)
-   $e_i$ = sampling error and is specified as a latent variable
-   $v_i$ = sampling variance across study and is fixed to individual study or subject.
:::

::: callout-note
### Work-flow

-   specifying the model using lavaan syntax
-   the proposed model is converted to RAM specification using lavaan2RAM function (the std.lv =False because we dont want to standardize the sampling variance to 1)
    -   RAM = Reticular Action Model
-   the data is fitted using sem function
-   model is displayed using the plot function
:::

::: callout-note
### model specification fixed-effect

-   Latent variable
    -   $e_i$ =\~ 1\*$y_i$
    -   Declares $e_i$ as a latent variable with $y_i$ as its sole observed indicator.
    -   The factor loading (regression coefficient) of $y_i$ on $e_i$ is fixed to 1, scaling the latent variable $e_i$ to the same units as $y_i$.
-   Mean Structure:
    -   $y_i$ \~ $\mu$\*1
    -   Specifies that the mean of $y_i$ is explicitly modeled as mu.
    -   This implies $\mu$ is an estimated parameter representing the average value of $y_i$
-   Residual Variance of Observed Variable:
    -   $y_i$ \~\~ 0\*$y_i$
    -   Sets the residual (error) variance of $y_i$ to 0.
    -   This means all variance in $y_i$ is fully explained by the latent variable $e_i$ (no measurement error).
    -   $y_i$ is treated as a perfect indicator of $e_i$.
-   Variance of Latent Variable:
    -   $e_i$ \~\~ data.vi\*$e_i$
    -   Assigns the variance of ei to values provided in the dataset’s vi column.
    -   Instead of estimating the latent variable’s variance, it uses case-specific pre-defined values (e.g., known sampling variances in meta-analysis or heteroscedasticity adjustments).
    -   This allows the model to incorporate external information about variability (e.g., study-specific variances in meta-analytic models)
:::

**R Code**:

```{r}
## Prepare a sample data
dat1 <- data.frame(yi=Jaramillo05$r, 
                   vi=Jaramillo05$r_v,
                   xi=scale(Jaramillo05$IDV, scale=FALSE),
                   zi=scale(Jaramillo05$Citations, scale=TRUE))

```

## Models without any external variables

### Fixed-effect model

```{r}
## Fixed-effect model with a latent variable ei
m1a <- "ei =~ 1*yi           ## e as a latent variable and yi is its indicator
        yi ~ mu*1            ## Mean(yi) = mu
        yi ~~ 0*yi           ## Set the error variance of yi at 0
        ei ~~ data.vi*ei     ## Var(yi) = vi, which is assigned from data
"

## Fixed-effects model without specifying ei
m1b <- "yi ~ mu*1            ## Mean(yi) = mu
        yi ~~ data.vi*yi     ## Var(yi) = vi, which is assigned from data
"


## Convert the lavaan model to RAM specification
## obs.variables = "yi": "yi" is the observed variable
ram1 <- lavaan2RAM(m1a, obs.variables = "yi", std.lv = FALSE)

## Display the RAM model
#ram1

## Get the model-implied mean and variance structures
#impliedS(ram1)

## Fit the model to the data
fit1 <- sem("Fixed", RAM = ram1, data = dat1)
summary(fit1)

## Display the model with results using the semPlot package
## nDigits=4: Display 4 digits
plot(fit1, color = "green", sizeInt = 7, nDigits = 4)
```

::: callout-note
### Result

-   $\hat{\mu}$ = .1938, \[.1782, .2093\]
:::

------------------------------------------------------------------------

## **Random-Effects Model**

-   Population effect-sizes can be different for individual studies.
-   The differences in the observed effect-sizes are due to a combination random (sampling) error with a variance of v (the difference between the observed and population effect-size) and the True difference (or variance) that is., the difference between the studies effect sizes.
-   Findings from the meta analysis can be generalized beyond the studies in the meta-analysis.

### Model Specification

**Equation**: $$
y_i = \mu + u_i + e_i, \quad u_i \sim N(0, \tau^2), \quad e_i \sim N(0, v_i)
$$

::: callout-note
### Model Specification Random Effect

-   Latent Variable (Random Effect)
    -   ui =\~ 1\*yi
    -   Declares ui as a latent variable (random effect) with yi as its observed indicator.
    -   The factor loading is fixed to 1, linking ui and yi on the same scale.
-   Variance of the Random Effect:
    -   ui \~\~ tau2\*ui
    -   Specifies that the variance of ui (denoted as tau2) is estimated from the data.
    -   tau2 represents the between-group or between-study heterogeneity (e.g., between-study variance in meta-analysis).
-   Mean Structure:
    -   yi \~ mu\*1
    -   Models the mean of yi as mu, an estimated parameter.
    -   mu represents the overall average effect (e.g., the pooled effect in meta-analysis).
-   Residual Variance of Observed Variable:
    -   yi \~\~ data.vi\*yi
    -   Sets the residual variance of yi to pre-specified values from the dataset’s vi column.
    -   This represents known measurement error variances (e.g., sampling variances in meta-analysis).
:::

**R Code**:

### Random-effects model

```{r}
## Random-effects model
m2 <- "ui =~ 1*yi          ## ui is the random effect
       ui ~~ tau2*ui       ## Var(ui) = tau2
       yi ~ mu*1           ## Mean(yi) = mu
       yi ~~ data.vi*yi    ## Var(yi) = vi, which is assigned from data
"

## Display the conceptual model
#plot(m2, color="yellow", sizeInt = 7)

## Convert the lavaan model to RAM specification
## std.lv = FALSE: Ensure the latent variable is not standardized
ram2 <- lavaan2RAM(m2, obs.variables = "yi", std.lv = FALSE)

## Get the model-implied mean and variance structures
#impliedS(ram2)

## intervals.type = "LB": Request likelihood-based CI
fit2 <- sem("Random", RAM = ram2, data = dat1, intervals.type = "LB")
summary(fit2)

## Display the model with results
plot(fit2, color="green", sizeInt = 7, nDigits = 4)
```

::: callout-note
### Result

-   $\hat{\mu}$ = .1866, \[.1479, .2251\] and $\hat{\tau}$ = 0.0170, \[0.0106, 0.0276\]
:::

```{r}
anova(fit1, fit2)
```

------------------------------------------------------------------------

## **Summary** {.unnumbered}

| Model | Equation | SEM Component | R Function |
|----|----|----|----|
| Fixed-Effect | $y_i = \beta_F + e_i$ | No latent variables | `meta(y, v)` |
| Random-Effects | $y_i = \beta_R + u_i + e_i$ | Latent $f_i \sim N(\beta_R, \tau^2)$ | `meta(y, v)` |
| Mixed-Effects | $y_i = \beta_0 + \beta_1 x_i + u_i + e_i$ | Latent $f_i \sim N(\beta_0 + \beta_1 x_i, \tau^2)$ | `meta(y, v, x)` |

::: callout-important
### Advantages of SEM-Based Meta-Analysis

-   **Flexibility**: Extend to multivariate/multilevel models.\
-   **Precision**: Directly model heterogeneity as latent variance.\
-   **Robustness**: Integrate with SEM's estimation tools (e.g., FIML, constraints).\
:::
