---
title: "W6 In-class Exercise: Ch7"
output: html_document
date: "2025-02-17"
---

# Ready

If you haven't installed the following packages, run the codes below:

```{r eval = FALSE}
install.packages("dplyr")
install.packages("metafor")
```

Load the packages

```{r}
library(dplyr)
library(metafor)
```

# `metafor` R package

The `metafor` package includes the \href{https://wviechtb.github.io/metafor/reference/escalc.html}{`escalc` function} for computing a wide range of ES estimates. See its online document for details.

-   NOTE: \texttt{escalc} uses a slightly different approximation for the sampling variance of $g$. You can change the variance estimator using the \texttt{vtype} argument.

# Ch 7. Exercise 1

```{r}
# Read data 
ex1dat <- read.csv("Ex1_dat.csv")

```

## Hand calculation:

```{r}
m_ctr <- mean(ex1dat$Control)
m_trt <- mean(ex1dat$Treatment)

v_ctr <- var(ex1dat$Control)
v_trt <- var(ex1dat$Treatment)

```

## Use `escalc`:

$SMD = \frac{m1i - m2i}{s_p}$, we define `m1i` = Treatment group's mean, `m2i` = Control group's mean.

```{r}
escalc(measure = "SMD", 
       m1i = mean(ex1dat$Treatment), sd1i = sd(ex1dat$Treatment), n1i = 8,
       m2i = mean(ex1dat$Control), sd2i = sd(ex1dat$Control), n2i = 8, 
       var.names = c("g", "v_g"))
```

# Ch 7. Exercise 3

```{r}
ex3dat <- read.csv("deltas.csv")
```

## Hand caluclation

```{r}

```

## Use metafor

```{r}

```
